Fractional Types

### 2016-12-18 typespp.tex
op sem

\begin{definition}[Groupoid Cardinality~\cite{2009arXiv0908.4305B}]
  The cardinality of a groupoid $G$ is the (positive) real number:
  \[
    |G| = \sum_{[x]} \frac{1}{|\textsf{Aut}(x)|}
  \]
  provided the sum converges. The summation is over \emph{isomorphism classes}
  $[x]$ of objects $x$ and $|\textsf{Aut}(x)|$ is the number of \emph{distinct}
  automorphisms of $x$.
\end{definition}

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite{looping}, that identify the two groupoids above.
Although the notion of equivalence we use in the next section appears novel, it
is related ``in spirit'' to Morita equivalence and we find it useful to give a
small example. We will verify that the product of the groupoid
$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
and one trivial automorphism. To avoid clutter we assume the groupoids are
strict, e.g., that the groupoid $\order{\permfive}$ has three elements (instead
of 3 isomorphism classes of elements). The definition of \emph{Morita
  equivalence} in this case reduces to to finding a \emph{Morita morphism} from
$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$. This is
a functor $\psi$ from the unit groupoid to the product groupoid, say $G$, with
an additional condition that the commutative diagram in figure~\ref{fig:morita}
is a \emph{pullback}. The intuition is that the codomain consists of three
objects with associated morphisms, each of them counting as a ``third.'' The
functor from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$ must include an arbitrary
choice of ``which third'' to target and the universality condition of the
pullback ensures that the choice is ultimately insignificant as it leads to the
same result as every other possible functor. To summarize the equivalence
between the two groupoids requires a \emph{local} map that chooses a target and
a \emph{global} condition ensuring that the choice ultimately annhilates. The
challenge we are facing is to turn this idea into an operational semantics that
somehow combines the local and global aspects of the equivalence.

\begin{figure}[h]
  \centering
  \begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G_1 \arrow[d, "{(s,t)}"] \\
    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
    & G_0 \times G_0
  \end{tikzcd}
  \caption{
    \label{fig:morita}
    Pullback diagram for a Morita morphism,
    $\order{\idiso} \rightarrow \order{\permfive} \circledast
    \iorder{\permfive}$
  }
\end{figure}

@@

@@

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Let the combinator $c$ to be of order greater than or
equal to 2, i.e., the type $\order{c}$ has at least two distinct values $c^i$
and $c^j$ and the type $\iorder{c}$ has at least two distinct values $1/c^i$ and
$1/c^j$. The use of $\eta+$ has then at least four choices; as discussed at the
end of Sec.~\ref{sub:products}, $\eta+$ will internalize some arbitrary but
fixed choice and might for example generate the pair
$\langle c^i , 1/c^i \rangle$. No matter what choice is internalized in $\eta+$,
there is at least one different incoming value $c^j$ of type $\order{c}$. In
that case, the combinator $\epsilon-$ encounters $\langle 1/c^i , c^j
\rangle$. If our operational semantics allows $\epsilon-$ to annihilate
arbitrary pairs, the net effect of the circuit would be to map $c^j$ to $c^i$
for arbitrary $i$ and $j$.

\begin{figure}[bht]
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
  \node at (0.5,1) {\footnotesize{$\unitetl/$}};
  \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (0,1);
  \path (1,1.75) edge node[above] {\footnotesize{$\order{\idiso}$}} (1.6,1.75);
  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
  \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,0.25);
  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
  \node at (2.1,1.9) {\footnotesize{$\eta+$}};
  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
  \path (2.6,2.6) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,2.6);
  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
  \path (2.6,1.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.25);
  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
  \node at (4.75,1.5) {$\footnotesize{\assocrt/}$};
  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
  \path (5.5,1.25) edge node[above] {$\footnotesize{1/\#c}$} (7,1.25);
  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
  \path (5.5,2.6) edge node[above] {\footnotesize{$\#c$}} (9,2.6);
  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
  \node at (7.5,0.8) {\footnotesize{$\epsilon-$}};
  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
  \path (8,1) edge node[above] {\footnotesize{$\order{\idiso}$}} (9,1);
  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
  \path (5.5,0.2) edge node[above] {\footnotesize{$\#c$}} (7,0.2);
  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
  \node at (9.5,1.8) {\footnotesize{$\unitetr/$}};
  \path (10,1.8) edge node [above] {\footnotesize{$\#c$}} (11,1.8);
\end{tikzpicture}
\caption{\label{fig:zigzag} Zig-zag}
\end{figure}

@@

The operational definition of synchr⋆ might seem at odds with our claim that it is related to associativity while it looks like commutativity.  The way to understand it is as follows: if indeed all values were indepedendent, it would be associativity.  However, in this case, $p // q$ is really two \emph{tangled} values, which we do not have access to.  To keep these synchronized and yet to achieve the given type, the only choice (operationally) is to swap.  This is where the ``action at a distance'' occurs.

>>>>

\begin{definition}[Groupoid cardinality~\cite{2009arXiv0908.4305B}]
  The cardinality of a groupoid $G$ is the (positive) real number:
  \[
    |G| = \sum_{[x]} \frac{1}{|\textsf{Aut}(x)|}
  \]
  provided the sum converges. The summation is over \emph{isomorphism classes}
  $[x]$ of objects $x$ and $|\textsf{Aut}(x)|$ is the number of \emph{distinct}
  automorphisms of $x$.
\end{definition}

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite{looping}, that identify the two groupoids above.
Although the notion of equivalence we use in the next section appears novel, it
is related ``in spirit'' to Morita equivalence and we find it useful to give a
small example. We will verify that the product of the groupoid
$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
and one trivial automorphism. To avoid clutter we assume the groupoids are
strict, e.g., that the groupoid $\order{\permfive}$ has three elements (instead
of 3 isomorphism classes of elements). The definition of \emph{Morita
  equivalence} in this case reduces to to finding a \emph{Morita morphism} from
$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$. This is
a functor $\psi$ from the unit groupoid to the product groupoid, say $G$, with
an additional condition that the commutative diagram in Fig.~\ref{fig:morita} is
a \emph{pullback}. The intuition is that the codomain consists of three objects
with associated morphisms, each of them counting as a ``third.'' The functor
from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor. To summarize the
equivalence between the two groupoids requires a \emph{local} map that chooses a
target and a \emph{global} condition ensuring that the choice ultimately
annihilates. The challenge we are facing is to turn this idea into an operational
semantics that somehow combines the local and global aspects of the equivalence.

\begin{figure}[t]
  \centering
  \begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G_1 \arrow[d, "{(s,t)}"] \\
    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
    & G_0 \times G_0
  \end{tikzcd}
  \caption{
    \label{fig:morita}
    Pullback diagram for a Morita morphism,
    $\order{\idiso} \rightarrow \order{\permfive} \circledast
    \iorder{\permfive}$
  }
\end{figure}

@@

\begin{figure}[t]
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
  \node at (0.5,1) {\footnotesize{$\unitetl/$}};
  \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (0,1);
  \path (1,1.75) edge node[above] {\footnotesize{$\order{\idiso}$}} (1.6,1.75);
  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
  \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,0.25);
  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
  \node at (2.1,1.9) {\footnotesize{$\eta+$}};
  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
  \path (2.6,2.6) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,2.6);
  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
  \path (2.6,1.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.25);
  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
  \node at (4.75,1.5) {$\footnotesize{\assocrt/}$};
  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
  \path (5.5,1.25) edge node[above] {$\footnotesize{1/\#c}$} (7,1.25);
  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
  \path (5.5,2.6) edge node[above] {\footnotesize{$\#c$}} (9,2.6);
  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
  \node at (7.5,0.8) {\footnotesize{$\epsilon-$}};
  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
  \path (8,1) edge node[above] {\footnotesize{$\order{\idiso}$}} (9,1);
  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
  \path (5.5,0.2) edge node[above] {\footnotesize{$\#c$}} (7,0.2);
  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
  \node at (9.5,1.8) {\footnotesize{$\unitetr/$}};
  \path (10,1.8) edge node [above] {\footnotesize{$\#c$}} (11,1.8);
\end{tikzpicture}
\caption{\label{fig:zigzag} Zig-zag}
\end{figure}

Each combinator $\rho$ has an inverse.
\caption{$\Pi^/$ fractional combinators with pointed types (dependent version)}
\label{pifracd:comb}
\end{figure}

@@

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to modify
the operational semantics to use a mechanism to express a computational effect
which enables spatially separated parts of the program to communicate in a way
that is reminiscent of entanglement in quantum mechanics. Possibilities to
realize this communication are global reference cells, backtracking, logical
variables with unification. We choose a more abstract and enlightening approach
that encodes the required dependency in dataflow constraints using dependent
types.

\begin{definition}[Dependent $\Pi^/$ with pointed types]
\label{def:pifracd}
The syntax of dependent $\Pi^/$ with pointed types differs from
Def.~\ref{def:pifrac} as follows:
\[\begin{array}{lrcl}
(\textrm{Pointed Fractional Types}) & \mathbb{T_\bullet} &::=& [\mathbb{T},v] \\
 (\textrm{Pointed /-combinators}) & \rho_\bullet &:&
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2] ~::=
  [\textit{see Fig.~\ref{pifracd:comb}}]
\end{array}\]
\end{definition}

@@

The operational definition of synchr⋆ might seem at odds with our claim that it is related to associativity while it looks like commutativity.  The way to understand it is as follows: if indeed all values were independent, it would be associativity.  However, in this case, $p // q$ is really two \emph{tangled} values, which we do not have access to.  To keep these synchronized and yet to achieve the given type, the only choice (operationally) is to swap.  This is where the ``action at a distance'' occurs.
